Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2022-02-04 manu-draft.Rmd
Minor edits - will continue on Monday
Trying to find a good estimate of total holding capacity of stacks, 2020 DEP presentation to RPC give a total of 1.3 billion gallons capacity for all of PP, so that's about 4.9 billion liters... - https://drive.google.com/drive/u/0/folders/1NC47EqfqivSDJ8GP_0H6AbUv3GO5chw_ - slide 16, p 8

Legacy mining facilities pose significant risks to aquatic resources.  From March 30th to April 9th, 2021, 814 million liters of phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA). This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days. An initial phytoplankton bloom (non-harmful diatoms) was first observed in April. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrasses had minimal changes over the study period.  By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to external nitrogen loads into Tampa Bay.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. The production of phosphate fertilizer generates large amounts of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O) that is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.  Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].  In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than natural surface waters [@Burnett01;@Beck18b]. Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change and seasonal stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site. Numerous studies have documented the environmental and human health risks associated with these stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

>>>>

Legacy mining facilities pose significant risks to aquatic resources.  From March 30th to April 9th, 2021, 814 million liters of phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA). This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days. An initial phytoplankton bloom (non-harmful diatoms) was first observed in April. Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*. Reported fish kills tracked *K. brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish. Seagrasses had minimal changes over the study period.  By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to abnormally high and rapidly delivered nitrogen loads into Tampa Bay.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. The production of phosphate fertilizer generates large amounts of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O) that is typically stored on-site in large earthen stacks (gypstacks) capable of holding hundreds of millions of liters of process water.  Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].  In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than natural surface waters [@Burnett01;@Beck18b]. Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change and seasonal stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site. Numerous studies have documented the environmental and human health risks associated with these stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
